State of the Practice for Lattice Boltzmann Method Software

### 2022-01-31 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Incorporating additional information into the comparison b/w our ranking and the community's ranking

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice}, such as
reflective and non-reflective boundaries. LBM considers the behavior of a
collection of particles as a single unit at the mesoscopic scale, which lies
between the nanoscopic and microscopic scales. LBM solvers predict the
positional probability of a collection of particles moving through a lattice
structure following a two step process: i) streaming, where the particles move
along the lattice via links; and, ii) colliding, where energy and momentum is
transferred among particles that collide \citep{bao2011lattice}. As an example
of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds
numbers. \wss{Zahra can likely provide a better example.}  LBM has several
advantages over conventional CFD methods, including a simple calculation
procedure, improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modeling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Since their time is valuable, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for existing LBM libraries,
the time is right to assess the state of the practice.  The goal of this report
is to analyze the current state of LBM software development and compare it to
the development practices employed for research software in general.  We want to
highlight success stories that LBM developers can share amongst their
community, and with the broader research software community, while at the same time watching for areas for potential future improvement.

@@

Filtering by scope, usage, and age decreased the size of the list to 23
packages. Many of the 22 packages that were removed could not be tested as there
was no installation guide, they were incomplete, source code was not publicly
available, a license was needed, or the project was out of scope or not up to a
standard that would support incorporating them into this study. These eliminated
software packages are listed in the Appendix of \citet{Michalski2021}. Of the
remaining 23 packages that were studied, some were kept on the list despite
being marked as dead due to their prevalence on authoritative lists on LBM
software and due to their surface excellence.  After the initial measurement
exercise, which took place in mid 2020, we added one more package in January
2022: Musubi.  Musubi did not come up in our initial search, but once we learned
of it, we felt that this report is incomplete without including it.

@@

Figure~\ref{Fig_Reusabilty} shows the reusability ranking of the software
packages. Software packages with a high score have thousands of source code
files and API documentation. The highest scoring packages, ESPResSo and
waLBerla, have extensive functionality, including graphical visualizations and
non-LBM modeling. For this reason, a comparison with other software packages is
not on a level field. However, these packages do have an abundance of reusable
components. Nine of the top 10 ranked packages are noted as being alive.

@@

To address \rqref{RQ_CompareHQ2Popular}, Table~\ref{repometrics} compares our
LBM software package rankings against their popularity in the research
community, as estimated by repository stars and watches. Nine packages do not
use GitHub, so they do not have a measure of repository stars. Looking at the
repository stars of the other 15 packages, we can observe a pattern where
packages that have been highly ranked by our assessment tend to have more stars
than lower ranked packages. Our best ranked package (ESPResSo) has the second
most stars, while our ninth ranked package (Sailfish) has the highest number of
stars. The same correlation is observed in the repository watch column, although
this column contains less data, since two of the packages (Palabos, waLBerla)
use GitLab, which does not track watches. Consistent with the star measure, our
ninth ranked package (Sailfish) has the highest number of watches and our best
ranked package (ESPResSo) has the second most watches. Packages designated as
lower quality often do not use GitHub or GitLab, or have only a few stars/watches.
Although the details of the rankings differ, our measures show that following
best practices correlates with increased popularity in the scientific community.

@@

Although our ranking and the estimate of popularity are consistent, trying to
make more specific conclusions from this data would be problematic.  To start
with, using stars and watches may not accurately measure popularity or
necessarily represent the perceptions of the community. Even if the stars and
watches are a good proxy for popularity, they are an unfair measure since
repositories vary in age. Older packages have had more time to accumulate stars
and watches. For instance, HemeLB only has 12 stars, but it is young because (at
the time of measurement) HemeLB had recently moved from a different GitHub
repository. From the other side, our estimate of ranking by best practices is
also an approximation.  Besides our use of shallow measures, the assumption of
equal weighting between all qualities to create our overall ranking
(Section~\ref{Sec_OverallQuality}) does not represent reality.  For real
projects some qualities are going to be more important than others.

Although our ranking and the estimate of the community's ranking are not perfect
measures, they do suggest a correlation between best practices and popularity.
We do not know which comes first, the use of best practices or popularity, but
we do know that the top ranked packages share common practices.  The next
sections will explore how the practices of the LBM community compare to the
broader research software community.  We will also investigate the practices
from the top projects that others within the LBM community, and within the
broader research software community, can potentially adopt.

@@

A few obstacles related to users were found. Several developers noted that users
sometimes try to use incorrect LBM method combinations to solve their problems.
Furthermore, some users think that the packages will work out of the box to
solve their cases, while the packages require both a good understanding of CFD
and an understanding of the requirements for formulating problems in the
individual packages, which can be a significant endeavor. These software
packages are not like commercial software packages. They are generally set up to
solve specific research problems, and are often primarily used by their
developers. While they are modifiable to solve similar problems, these
modifications are not trivial. Better documentation, with attention on
traceability, and automatic code generation are suggested when designing
software for change, and would help with these modifications. \wss{Should decide
if want to introduce ideas for improving within the presentation of the pain
point, or separately?}  So far this problem of on-boarding new users has been
addressed by updating the documentation to better inform users of the underlying
LBM theory and package requirements. Similar issues with LBM parameters were
noted by another developer. Updating the user interface to better explain
theoretical principles, as well as test user input for compatibility, was the
implemented solution. As noted above, sometimes frequently asked questions on
the underlying theory and on how to use the software are answered in the
documentation.  The interviewed developer of ESPResSo commented that parts of
their package's source code had been refactored to Python to help address
usability issues. Python was perceived as a much more usable language, and it
would be easy for future users and developers to learn and understand the source
code.

@@

Some software packages, like Palabos, provide validation benchmarks for their
core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
intent of these benchmarks is to showcase the validity and usefulness of the
package to stimulate the development of third-user extensions. The Palabos
package identifies as a development framework for modeling problems in various
CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
Palabos - others could emmulate.  Goal of this exercise is for people in the
community to learn from each other, and potentially from our advice too.}

@@

Furthermore, design documentation was measured to be a ``less common'' artifact
in this assessment, but it is probable that such documentation is part of all
LBM packages. After all, developing SCS is not a trivial endeavor. It is likely
that many packages have such documentation but did not make it public, and due
to this the measured data is not a true reflection of software package quality.

@@

>>>>

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice}, such as
reflective and non-reflective boundaries. LBM considers the behaviour of a
collection of particles as a single unit at the mesoscopic scale, which lies
between the nanoscopic and microscopic scales. LBM solvers predict the
positional probability of a collection of particles moving through a lattice
structure following a two step process: i) streaming, where the particles move
along the lattice via links; and, ii) colliding, where energy and momentum is
transferred among particles that collide \citep{bao2011lattice}. As an example
of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds
numbers. \wss{Zahra can likely provide a better example.}  LBM has several
advantages over conventional CFD methods, including a simple calculation
procedure, improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Since their time is valuable, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for existing LBM libraries,
the time is right to assess the state of the practice.  The goal of this report
is to analyze the current state of LBM software development and compare it to
the development practices employed for research software in general.  We want to
highlight success stories that LBM developers can share amongst their
community, and with the broader research software community, while at the same time watching for areas for potential future improvement.

@@

Filtering by scope, usage, and age decreased the size of the list to 23
packages. Many of the 22 packages that were removed could not be tested as there
was no installation guide, they were incomplete, source code was not publicly
available, a license was needed, or the project was out of scope or not up to a
standard that would support incorporating them into this study. These eliminated
software packages are listed in the Appendix of \citet{Michalski2021}. Of the
remaining 23 packages that were studied, some were kept on the list despite
being marked as dead due to their prevalence on authoritative lists on LBM
software and due to their surface excellence.  After the initial measurement
exercise, which took place in mid 2020, we added one more package in January
2022: Musubi.  Musubi did not come up in our initial search, but once we learned
of it, we felt that this report is incomplete without including it.  With the
addition of Musubi we did not update the manual collection of data for the
previous packages, due to time constraints, but we did redo all data collection
that was automated by tools.

@@

Figure~\ref{Fig_Reusabilty} shows the reusability ranking of the software
packages. Software packages with a high score have thousands of source code
files and API documentation. The highest scoring packages, ESPResSo and
waLBerla, have extensive functionality, including graphical visualizations and
non-LBM modelling. For this reason, a comparison with other software packages is
not on a level field. However, these packages do have an abundance of reusable
components. Nine of the top 10 ranked packages are noted as being alive.

@@

To address \rqref{RQ_CompareHQ2Popular}, Table~\ref{repometrics} compares our
LBM software package rankings against their popularity in the research
community, as estimated by repository stars and watches. Nine packages do not
use GitHub, so they do not have a measure of repository stars. Looking at the
repository stars of the other 15 packages, we can observe a pattern where
packages that have been highly ranked by our assessment tend to have more stars
than lower ranked packages. Our best ranked package (ESPResSo) has the second
most stars, while our ninth ranked package (Sailfish) has the highest number of
stars. The same correlation is observed in the repository watch column, although
this column contains less data, since two of the packages (Palabos, waLBerla)
use GitLab, which does not track watches. Consistent with the star measure, our
ninth ranked package (Sailfish) has the highest number of watches and our best
ranked package (ESPResSo) has the second most watches. Packages designated as
lower quality often do not use GitHub or GitLab, or have only a few
stars/watches. Although the details of the rankings differ, our measures show
that following best practices tends to correlate with increased popularity in
the scientific community.

Notable exceptions to this trend occur for three projects that rank high by star
count, but relatively low in the ``follow best practices'' ranking: TCLB (star
rank 3, our rank 12), lettuce (star rank 4, our rank 14) and ESPResSo++ (star
rank 5, our rank 15).  Part of the discrepancy is because all projects are
ranked by our process, but not all projects are measured by stars.  Also, our
process equally weights all qualities (Section~\ref{Sec_OverallQuality}), but
for real users some qualities will be more important than others. Discrepancies
between our measures and the star measures may also be due to inaccuracy with
using stars to approximate popularity, since stars and watches tend to increase
over time. Older packages (like ESPResSo++ and TCLB) have had more time to
accumulate stars and watches. Young projects have less time to accumulate stars,
like HemeLB, which only 12 stars (at the time of measurement) having recently
moved from a different GitHub repository.  Lettuce is a young project (released
in 2019) with a relatively high number of stars.  The discrepancy in rankings
here may be due to a mismatch between the time when lettuce was manually
measured by our template and when the stars were counted.  The late addition of
Musubi meant a large gap between when projects were manually measured and the
time when our automated data (like stars) was collected.

@@

Although our ranking and the estimate of the community's ranking are not perfect
measures, they do suggest a correlation between best practices and popularity.
We do not know which comes first, the use of best practices or popularity, but
we do know that the top ranked packages tend to incorporate best practices.  The
next sections will explore how the practices of the LBM community compare to the
broader research software community.  We will also investigate the practices
from the top projects that others within the LBM community, and within the
broader research software community, can potentially adopt.

@@

A few obstacles related to users were found. Several developers noted that users
sometimes try to use incorrect LBM method combinations to solve their problems.
Furthermore, some users think that the packages will work out of the box to
solve their cases, while the packages require both a good understanding of CFD
and an understanding of the requirements for formulating problems in the
individual packages, which can be a significant endeavour. These software
packages are not like commercial software packages. They are generally set up to
solve specific research problems, and are often primarily used by their
developers. While they are modifiable to solve similar problems, these
modifications are not trivial. Better documentation, with attention on
traceability, and automatic code generation are suggested when designing
software for change, and would help with these modifications. \wss{Should decide
if want to introduce ideas for improving within the presentation of the pain
point, or separately?}  So far this problem of on-boarding new users has been
addressed by updating the documentation to better inform users of the underlying
LBM theory and package requirements. Similar issues with LBM parameters were
noted by another developer. Updating the user interface to better explain
theoretical principles, as well as test user input for compatibility, was the
implemented solution. As noted above, sometimes frequently asked questions on
the underlying theory and on how to use the software are answered in the
documentation.  The interviewed developer of ESPResSo commented that parts of
their package's source code had been refactored to Python to help address
usability issues. Python was perceived as a much more usable language, and it
would be easy for future users and developers to learn and understand the source
code.

@@

Some software packages, like Palabos, provide validation benchmarks for their
core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
intent of these benchmarks is to showcase the validity and usefulness of the
package to stimulate the development of third-user extensions. The Palabos
package identifies as a development framework for modelling problems in various
CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
Palabos - others could emmulate.  Goal of this exercise is for people in the
community to learn from each other, and potentially from our advice too.}

@@

Furthermore, design documentation was measured to be a ``less common'' artifact
in this assessment, but it is probable that such documentation is part of all
LBM packages. After all, developing SCS is not a trivial endeavour. It is likely
that many packages have such documentation but did not make it public, and due
to this the measured data is not a true reflection of software package quality.

@@

Threat to validity - the gap between measures of everything else and Musubi
(Section~\ref{identifysoftware}), and then the gap between the manual measures
and the automated measures (Section~\ref{repmetrics}).
